---
title: "Solving ODEs"
---

This tutorial describes how to implement and solve an ODE system in R. 
The base R installation has no ODE solving capacities, hence a package has to 
be used. Here, we will use the most frequently used one, the `deSolve` package.

First install (if not done yet) and load the package:

```{r, eval = FALSE}
install.packages("deSolve")
```
```{r}
library("deSolve")
```

## Syntax of function `ode`

Function `ode` solves ODE systems. It has the following syntax:

Its arguments are:

* `y`, the initial values
* `times`, 
* `func`, the right-hand side (an R function),
* `parms`, a parameter vector used in function `func` (useful e.g. for optimization)

Additionally, an ODE solver can be defined, but there is often no need to 
customize the default solver "lsoda", which solves both non-stiff and stiff ODE
systems efficiently.

### Syntax of argument `func`

Argument `func` needs to be a function with the following syntax (see tutorial
"Defining a function" for general information about how to define functions
yourself):

```{r}
func <- function(t, y, p) {
    #
    # computation
    #
    list(dy) # return value as a one-element list
}
```

It must have three arguments (in that order):

* the current time `t` during solving the ODE 
  (must be specified even for time-independent systems, where `t` will be unused)
* `y`, the state vector (a numeric vector)
* `p`, the parameter vector

The ODE right-hand side must return a (length one) *list containing the* vector of 
derivatives `dy`, **not** the vector of derivatives itself!

### Ouput of `ode`

Function `ode` returns a matrix, below denoted `out`, containing 

* the output times as first column,
* the different states of the ODE system as subsequent columns

If the initial condition `y` was named, so will be the columns of this matrix.

## Examples

### An easy example

Consider the ODE $y' = - \,y, y(0) = 1$ (exponential decay), 
to be solved during the time interval $[0,10]$:

```{r}
y0 <- c(y=1) # name this --> output named
t  <- 0:10
rhs <- function(t, y, p) {
    dy <- -y
    list(dy)
}
out <- ode(y = y0, times = t, func = rhs)
out
```

Here, there are no parameters, and argument `parms` of function `ode` can be 
left unspecified. In contrast, the third argument, `p`, of the ODE right-hand side 
function must be given, although that argument is unused.

### A more complex example

Now consider the ODE system
$$x' = -a\,x + b\,y,\quad y' = -c\,y + d\,x$$
with initial condition $x(0)=1$, $y(0)=2$, parameters $a=b=c=d=1$ and timespan
$[0,10]$:

```{r}
y0 <- c(x=1,y=2) # name this --> output named
t  <- 0:10
p <- c(a=1,b=1,c=1,d=1)
rhs2 <- function(t, Y, p) { # Y = (x,y)
    dx <- - p["a"] * Y["x"] + p["b"] * Y["y"]
    dy <- - p["c"] * Y["y"] + p["d"] * Y["x"]
    dY <- c(dx,dy)
    list(dY)
}
out <- ode(y = y0, times = t, func = rhs2, parms = p)
out
```

Here, to make the code easier to read, it is good practice to use named states 
and parameters and refer to their names within the right-hand side function.

#### Alternative implementation of the right-hand side

An alternative style of encoding the right-hand side (you can also find this in
the Examples section of the function help for `ode`) uses function `with`:

```{r, eval = FALSE}
rhs2 <- function(t, Y, p)  {
  with(as.list(c(p, Y)), {
    dx <- -a * x + b * y
    dy <- -c * y + d * x
    dY <- c(dx, dy)
    list(dY)
  })
}
```

This style makes the actual computations easier to read. However, function `with`
uses non-standard evaluation, which has its downsides, too, for example:

* `p` and `Y` are not allowed to have overlapping names
* error messages are more difficult to read
* a computational overhead
